Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (\emph{Sida fallax})

### 2023-09-06 ms/si.tex
initial commit all files

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

% make better hyphen for numeric ranges, \numrange{}{}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\addtokomafont{disposition}{\rmfamily}
\RedeclareSectionCommand[
  font=\normalfont\Large]{section}
\RedeclareSectionCommand[
  font=\normalfont\normalsize\bfseries]{subsection}
\RedeclareSectionCommand[
  font=\normalfont\normalsize\itshape]{subsubsection}
\RedeclareSectionCommand[
  font=\normalfont\normalsize]{paragraph}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (Sida fallax)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Amphistomy increases leaf photosynthesis more in coastal than
montane plants of Hawaiian ʻilima (\emph{Sida fallax})}
\author{}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, enhanced]}{\end{tcolorbox}}\fi

\begin{center}
Genevieve Triplett$^1$, Thomas N. Buckley$^2$, Christopher D. Muir$^{1,3,*}$
\end{center}

\(^1\) School of Life Sciences, University of Hawaiʻi Mānoa, Honolulu,
HI, USA 96822

\(^2\) School of Life Sciences, University of California, Davis, CA, USA
95616

\(^3\) Department of Botany, University of Wisconsin, Madison, WI, USA
53706

\(^*\) Corresponding author: Christopher D. Muir, cdmuir@wisc.edu

Manuscript received \_\_\_\_\_\_\_; revision accepted \_\_\_\_\_\_\_.

Running head: Amphistomy advantage in ʻilima

@@

\hypertarget{abstract}{%
\section{ABSTRACT ---}\label{abstract}}

\noindent \textbf{Premise of the study} The adaptive significance of
stomata on both upper and lower leaf surfaces, called amphistomy, is
unresolved. A widespread association between amphistomy and open, sunny
habitats suggests the adaptive benefit of amphistomy may be greatest in
these contexts, but this hypothesis has not been tested experimentally.
Understanding why amphistomy evolves can inform its potential as a
target for crop improvement and paleoenvironment reconstruction.

\textbf{Methods} We developed a new method to quantify ``amphistomy
advantage'', \(\mathrm{AA}\), as the ratio of photosynthesis in an
amphistomatous leaf to that of the same leaf with the same total
stomatal conductance but with gas exchange blocked through the upper
(adaxial) surface, which we term ʻpseudohypostomyʻ. We thus estimated
\(\mathrm{AA}\) and related physiological and anatomical traits in 12
populations, six coastal (open, sunny) and six montane (closed, shaded),
of the indigenous Hawaiian species ʻilima (\emph{Sida fallax}).

\textbf{Key results} Coastal ʻilima leaves benefit 4.58 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and that leaves with more
similar conductances at each surface benefit more than leaves with few
stomata on the upper surface.

\textbf{Conclusions} This is the first direct experimental evidence that
amphistomy \emph{per se} increases photosynthesis, consistent with the
hypothesis that parallel pathways through upper and lower mesophyll
increase the supply of CO\(_2\) to chloroplasts. The prevalence of
amphistomatous leaves in open, sunny habitats can partially be explained
the increased benefit of amphistomy in ʻsunʻ leaves, but the mechanistic
basis of this observation is an area for future research.

\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, leaf, \emph{Sida
fallax}, stomata

\hypertarget{introduction}{%
\section{INTRODUCTION ---}\label{introduction}}

Amphistomy, the presence of stomata on both lower and upper surfaces of
broad leaves, should increase carbon gain by reducing the average
diffusion pathlength between stomata and chloroplasts
(\autoref{fig:compare-circuit-diagrams}), yet paradoxically this
seemingly simple adaptation is uncommon in nature and we don't know why.
Understanding variation in stomatal traits like amphistomy is imperative
because these tiny pores play an outsized ecological role in the global
carbon and water cycles (Hetherington and Woodward, 2003; Berry et al.,
2010). A widely applicable, accurate representation of how stomata
mediate the relationship between CO\(_2\) gained through photosynthesis
and water lost through evapotranspiration is essential to predict future
climate using Earth Systems Models (Jarvis, 1976; Ball et al., 1987;
Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997). Optimality
models accurately predict the major cause of water loss, stomatal
conductance (\(g_\mathrm{sw}\)), by assuming plants maximize carbon gain
minus a cost of water (Cowan and Farquhar, 1977; Givnish, 1986; Medlyn
et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks et al., 2018;
Deans et al., 2020; Franklin et al., 2020; Wang et al., 2020; Harrison
et al., 2021). Despite the success of optimality modeling in predicting
\(g_\mathrm{sw}\), the same modeling approach has so far failed to
explain the rarity of amphistomatous leaves (Muir, 2019). \textbf{This
gap between theory and observations strongly implies that we remain
ignorant about some key benefits and costs associated with stomata.}

Where are amphistomatous leaves found and why aren't they more common?
Among terrestrial flowering plants, amphistomatous leaves are rarely
found on woody plants and shade-tolerant herbs, but they are common in
annual and perennial herbs from sunny habitats (Salisbury, 1928;
Parkhurst, 1978; Mott et al., 1982; Peat and Fitter, 1994; Gibson, 1996;
Jordan et al., 2014; Muir, 2015, 2018; Bucher et al., 2017). Even in
resupinate leaves where the abaxial surface faces up toward the sky,
stomata develop on the lower adaxial surface (Lyshede, 2002). Exceptions
to this general pattern include some arid woody plants which typically
have vertically oriented, isobilateral leaves (Wood, 1934; Jordan et
al., 2014; Boer et al., 2016; Drake et al., 2019) and
floating/amphibious leaves of aquatic plants (Kaul, 1976; Doll et al.,
2021). The dearth of amphistomatous leaves should be quite surprising
and has been described as one of the most important unsolved problems in
the study of leaf structure-function relations despite some recent
progress (Grubb, 1977, 2020).

Amphistomatous leaves should be common because, all else being equal, a
leaf with a given number of stomata per area could increase its
photosynthetic rate simply by apportioning approximately half its
stomata to each surface (Parkhurst, 1978; Gutschick, 1984a, b). The key
difference between a hypo- and amphistomatous leaf, holding all other
factors constant, is that an amphistomatous leaf has two parallel
diffusion paths through the internal airspace to any given chloroplast.
Those airspaces pose a resistance for CO\(_2\) diffusion, so CO\(_2\)
concentration drops as it approaches chloroplasts. Shorter pathways mean
a smaller drop in CO\(_2\) concentration. Thus, chloroplasts in
amphistomatous leaves experience higher CO\(_2\) concentrations than in
hypostomatous leaves, thereby increasing photosynthesis. The airspace
resistance (or its inverse, the airspace conductance,
\(g_\mathrm{ias}\)) is rarely measured directly and there is
disagreement between empirical (Parkhurst and Mott, 1990; Morison et
al., 2005; Evans et al., 2009; Tomás et al., 2013; Earles et al., 2018;
Šantrůček et al., 2019; Nobel, 2020; Harwood et al., 2021; Márquez et
al., 2023) and theoretical models (Tholen and Zhu, 2011; Ho et al.,
2016; Théroux-Rancourt et al., 2021). The \(g_\mathrm{ias}\) in thin,
porous leaves may be so large as to be inconsequential given much lower
conductances for other components of the diffusion pathway, whereas the
\(g_\mathrm{ias}\) of thick leaves with little airspace may greatly
hinder CO\(_2\) diffusion to chloroplasts. Amphistomy should confer the
largest photosynthetic benefit in leaves with intrinsically low
\(g_\mathrm{ias}\).

Amphistomy should also enhance photosynthesis when leaf boundary layer
resistance is high, because apportioning total flux between two boundary
layers rather than one results in a smaller CO\(_2\) concentration drop
between the atmosphere and stomata. A similar effect has been validated
with a computer model and measurements for transpiration: amphistomatous
leaves lose somewhat more water for the same vapor pressure deficit and
total \(g_\mathrm{sw}\) (Foster and Smith, 1986), but the additional
carbon gain should be enough to offset this cost under most realistic
conditions (Muir, 2019).

The most promising adaptive hypothesis is that amphistomy is important
for maximizing photosynthetic rate under high light. Mott et al. (1982)
proposed that ``plants with a high photosynthetic capacity, living in
full-sun environments, and experiencing rapidly fluctuating or
continuously available soil water'' would benefit most, in terms of
increased carbon gain, from having amphistomatous leaves. As described
above, herbs from sunny habitats are often amphistomatous. Most
variation in stomatal density ratio (\(\mathrm{SR}\), the ratio of
stomatal density between the upper and lower surfaces) among species is
assumed to be genetic, but there is also putatively adaptive plasticity
in response to light. Leaves of \emph{Ambrosia cordifolia}, a desert
perennial herb, are hypostomatous under low light (photosynthetic photon
flux density, PPFD = 110 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
but develop \textasciitilde20\% of their stomata on the upper surface
under high light (1700 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
(Mott and Michaelson, 1991). Similarly, \emph{Solanum lycopersicum}
leaves are hypostomatous when grown in the shade but develop
\textasciitilde20\% of their stomata on the upper surface grown under
high light-intensity (Gay and Hurd, 1975). Adult leaves of
\emph{Eucalyptus globulus} are amphistomatous, but the proportion of
adaxial stomata increases from \textasciitilde10-20\% under low light to
\textasciitilde30-40\% under high light (James and Bell, 2001). In
summary, both genetic and plastic responses to evince a widespread
association between light and \(\mathrm{SR}\).

The association between high light and amphistomy suggests that `sun'
leaves have the most to gain in terms of increased photosynthesis from
having stomata on both surfaces, as Mott et al. (1982) hypothesized.
Parkhurst (1978) proposed quantifying this benefit as `amphistomy
advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
(see Materials and Methods). This hypothesis has never been tested
directly by comparing the photosynthetic rate of an amphistomatous leaf
to that of an otherwise identical hypostomatous leaf with the same total
stomatal conductance under the same conditions. We propose a
straightforward method to do this by experimentally creating a
pseudohypostomatous leaf with gas exchange blocked through the upper
surface (see Materials and Methods). One reason that sun leaves might
have greater \(\mathrm{AA}\) is that they are usually thicker or denser
(Poorter et al., 2019), resulting in lower \(g_\mathrm{ias}\)
(Parkhurst, 1978). A nonmutually exclusive hypothesis is that if sun
leaves have a stomatal density ratio closer to 0.5 (same density on each
leaf surface), this will confer a greater advantage than an
amphistomatous leaf with most stomata on one surface. In other words,
amphistomy doesn't make much difference if one leaf surface has very
open stomata on it. We therefore predict that sun leaves will have
greater \(\mathrm{AA}\) possibly because they have thicker leaves and/or
\(\mathrm{SR}\) closer to 0.5. We actually report
\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
that takes into account differences in both stomatal density and size
between surfaces.

The native flora of the Hawaiian archipelago is a excellent system to
test the relationship between light habitat and \(\mathrm{AA}\). Many
lineages have adapted to different light habitats after colonization and
leaf anatomical traits such as \(\mathrm{SR}\) and thickness vary within
and among closely related species. It is hypothesized that the common
ancestor in many Hawaiian clades was a ``weedy'' species with high
dispersal ability adapted to open habitats (Carlquist, 1966).
Colonization was followed by adaptive radiation into higher elevation,
montane, closed, forested habitats. Consequently, adaptation to sun and
shade is a common axis of phenotypic variation among Hawaiian plants
such as lobeliads (Givnish et al., 2004; Montgomery and Givnish, 2008;
Givnish et al., 2009; Givnish and Montgomery, 2014; Scoffoni et al.,
2015), \emph{Bidens} (Carlquist, 1966; Knope et al., 2020),
\emph{Scaevola} (Robichaux and Pearcy, 1984; McKown et al., 2016),
\emph{Euphorbia} (Sporck, 2011), and \emph{Plantago} (Dunbar-Co et al.,
2009).

Here we focus on variation within an indigenous plant species \emph{Sida
fallax} Walp. (Malvaceae), known in the Hawaiian language as ʻilima.
ʻIlima is found from sea level to elevations \(>1000\) mas on multiple
Hawaiian islands. Coastal populations are morphologically different from
montane populations (\autoref{fig:study-system}). Coastal regions of
Hawaiʻi are characterized by high sun exposure, warmer temperatures,
high winds, salinity, and variation in water availability. Coastal
populations of ʻilima tend to be short and prostrate which likely helps
them to withstand the windy environment (\autoref{fig:study-system}a).
The leaves of these populations are covered on both surfaces in dense,
soft hairs that give the leaves a silvery green appearance
(\autoref{fig:study-system}b), which helps mitigate water loss by
reflecting solar radiation, thereby lowering leaf temperature
(Ehleringer and Björkman, 1978). Montane regions, on the other hand,
provide very different challenges. Many other tall species grow on the
slopes of these wet mountainous regions, which makes light competition a
factor that plants may need to adapt to. Possibly due to this, montane
populations are erect and shrub- or tree-like, capable of growing meters
tall with strong, woody stems. These individuals have smooth, green
foliage with serrated edges. Montane populations exhibit traits that
help them to compete for light availability (Raunch et al.~1997). This
montane morphology is not found in \emph{S. fallax} populations on other
Pacific Islands (Pejhanmehr, 2022). The variable morphology between
coastal and montane populations has even led some scientists to argue
that it makes up a species complex, rather than a single species (Raunch
et al.~1997).

The objectives of our study are to test whether 1) `sun' leaves of
coastal ʻilima plants will have greater \(\mathrm{AA}\) than `shade'
leaves of montane plants; and if so, is this because 2a) coastal plants
have thicker leaves than montane plants and/or 2b) coastal plants have a
\(g_{\mathrm{smax,ratio}}\) closer to 0.5?

\begin{figure}[H]
  \includegraphics{../figures/study-system.pdf}
  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
  \label{fig:study-system}
\end{figure}

\hypertarget{methods}{%
\section{MATERIALS AND METHODS ---}\label{methods}}

\hypertarget{plant-sampling}{%
\subsection{Plant sampling ---}\label{plant-sampling}}

We identified 8 suitable natural populations of ʻilima on Oʻahu and 5 on
Hawaiʻi Island by consulting (Yorkston and Daehler, 2006) and citizen
scientist records on iNaturalist (Anon, 2022)
(\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites
that appeared to be cultivated. We visited sites between August and
November 2022. For logistical reasons, the sites on Hawaiʻi were sampled
in one three-day trip. We haphazardly sampled eight plants distributed
evenly between the highest and lowest elevation plants along a transect
at each site. For safety and conservation reasons, transects were along
a trail or road. We did not sample small individuals if there was risk
removing leaves would cause mortality. From each plant, we collected two
fully expanded leaves for traits measurements. We sampled stomatal
traits on all leaves; leaf thickness on one leaf from three randomly
selected plants per site; and, due to limited time, a single leaf from a
single plant at the middle of each transect for gas exchange
measurements.

\hypertarget{tbl-locations}{}
\begin{table}
\caption{\label{tbl-locations}ʻIlima study sites for this study. }\tabularnewline

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

% make better hyphen for numeric ranges, \numrange{}{}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2023 – Appendix S1}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

\begin{landscape}
\begin{longtable}{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{1.5cm}rrrrrrr}
\caption{Average traits values among ʻilima leaves at each site.
\(\mathrm{SD}_\text{abaxial}\) is the stomatal density per mm\(^2\) on
the abaxial (lower) surface; \(\mathrm{SD}_\text{adaxial}\) is the
stomatal density per mm\(^2\) on the adaxial (upper) surface;
\(\mathrm{GCL}_\text{abaxial}\) is the guard cell length in \(\mu\)m on
the abaxial (lower) surface; \(\mathrm{GCL}_\text{adaxial}\) is the
guard cell length in \(\mu\)m on the adaxial (upper) surface; Leaf
thickness is the length from upper cuticle to lower cuticle in \(\mu\)m;
\(A\) is the photosynthetic rate in
\(\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}\); \(g_\mathrm{sw}\)
is the stomatal conductance to water vapor in
\(\text{mol}~\text{m}^{-2}~\text{s}^{-1}\).}\tabularnewline

\toprule
Site & Island & Habitat & $\mathrm{SD}_\text{abaxial}$ & $\mathrm{SD}_\text{adaxial}$ & $\mathrm{GCL}_\text{abaxial}$ & $\mathrm{GCL}_\text{adaxial}$ & Leaf thickness & $A$ & $g_\text{sw}$\\
\midrule
\cellcolor{gray!6}{Kaloko-Honokōhau national historical park} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{412.99} & \cellcolor{gray!6}{92.31} & \cellcolor{gray!6}{14.73} & \cellcolor{gray!6}{20.32} & \cellcolor{gray!6}{189.60} & \cellcolor{gray!6}{17.40} & \cellcolor{gray!6}{0.123}\\
Puakō petroglyph park & Hawaiʻi & coastal & 310.12 & 52.55 & 13.35 & 20.94 & 194.20 & 17.85 & 0.134\\
\cellcolor{gray!6}{Kahuku Point} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{420.43} & \cellcolor{gray!6}{80.80} & \cellcolor{gray!6}{16.23} & \cellcolor{gray!6}{22.77} & \cellcolor{gray!6}{304.24} & \cellcolor{gray!6}{20.88} & \cellcolor{gray!6}{0.155}\\
Kaloko beach & Oʻahu & coastal & 400.27 & 109.03 & 15.69 & 23.26 & 399.54 & 31.28 & 0.314\\
\cellcolor{gray!6}{Kaʻena Point} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{370.14} & \cellcolor{gray!6}{120.34} & \cellcolor{gray!6}{17.61} & \cellcolor{gray!6}{22.89} & \cellcolor{gray!6}{295.82} & \cellcolor{gray!6}{34.77} & \cellcolor{gray!6}{0.336}\\
\addlinespace
Makapuʻu beach & Oʻahu & coastal & 408.28 & 23.18 & 15.99 & 22.03 & 249.71 & 32.87 & 0.349\\
\cellcolor{gray!6}{Hāloa ʻĀina} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{307.94} & \cellcolor{gray!6}{7.61} & \cellcolor{gray!6}{16.49} & \cellcolor{gray!6}{22.59} & \cellcolor{gray!6}{149.41} & \cellcolor{gray!6}{13.07} & \cellcolor{gray!6}{0.116}\\
Kaʻohe game management area & Hawaiʻi & montane & 270.24 & 13.07 & 14.87 & 21.51 & 183.72 & 12.94 & 0.130\\
\cellcolor{gray!6}{Koaiʻa tree sanctuary} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{318.58} & \cellcolor{gray!6}{9.55} & \cellcolor{gray!6}{13.89} & \cellcolor{gray!6}{22.86} & \cellcolor{gray!6}{138.89} & \cellcolor{gray!6}{27.88} & \cellcolor{gray!6}{0.358}\\
Hawaiʻi loa ridge & Oʻahu & montane & 329.25 & 115.56 & 17.83 & 21.76 & 205.97 & 21.95 & 0.215\\
\addlinespace
\cellcolor{gray!6}{Mauʻumae Ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{298.77} & \cellcolor{gray!6}{138.32} & \cellcolor{gray!6}{16.53} & \cellcolor{gray!6}{20.72} & \cellcolor{gray!6}{162.02} & \cellcolor{gray!6}{24.60} & \cellcolor{gray!6}{0.436}\\
Waʻahila ridge & Oʻahu & montane & 346.75 & 150.26 & 17.76 & 20.61 & 193.33 & 13.96 & 0.194\\
\bottomrule
\end{longtable}
\end{landscape}

@@

\begin{figure}
  \includegraphics{../figures/licor.pdf}
  \caption{See next page.}
  \label{fig:licor}
\end{figure}

\setcounter{figure}{\numexpr\value{figure}-1\relax}

\begin{figure}
  \caption{(Contunited from previous page) Individual $A \textendash g_\mathrm{sw}$ curves used to estimate $\mathrm{AA}$. For each leaf, one per site, we measured $A$ ($y$-axis) over a range of $g_\mathrm{sw}$ ($x$-axis) on the same leaf with two treatments: `amphi' (circles, solid line) leaves were untreated; `pseudohypo' (squares, dashed line) leaves had no conductance through the upper (adaxial) surface. In all coastal (orange) and montane (blue) leaves, we fit generalized additive models and 95\% confidence ribbons to estimate $\mathrm{AA}$ at a $g_\mathrm{sw}$ where the curves overlap (vertical black line). In leaves from Kaloko-Honokōhau national historical park and Puakō petroglyph park, we extrapolated sligtly beyond fitted curves because they did not quite overlap. $\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
  \label{fig:licor}
\end{figure}

\begin{figure}
  \includegraphics{../figures/habitat-Ags.pdf}
  \caption{The photsynthetic rate (left facet) and stomatal conductance to water vapor (right facet) of montane (blue) and coastal (orange) ʻilima leaves. Each point-interval is the median posterior estimate plus 95\% confidence interval of trait value for that habitat. Smaller points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants, one per site. $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
  \label{fig:habitat-Ags}
\end{figure}

\begin{figure}
  \includegraphics{../figures/habitat-gmaxratio.pdf}
  \caption{The $g_\mathrm{smax,ratio}$ ($y$-axis) of montane (blue) and coastal (orange) ʻilima leaves. Each point-interval is the median posterior estimate plus 95\% confidence interval of $g_\mathrm{smax,ratio}$ for that site. Sites are arranged by habitat and ascending $g_\mathrm{smax,ratio}$ within habitat. Smaller, transparent points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants. $g_\mathrm{smax,ratio}$, the ratio of anatomical maximum stomatal conductance to water vapor on the the adaxial surface over the total.}
  \label{fig:habitat-gmaxratio}
\end{figure}
